---
title: "Does Air Quality Affect Hospitalizations from Asthma in California?"
author: "Lewis White"
date: "2022-11-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading Packages

```{r}
#LOADING PACKAGES
library(tidyverse)
library(lubridate)
library(janitor)
library(here)
library(purrr)
library(fs)
library(sf)
library(tmap)
library(dplyr)
library(broom)
library(MASS)
library(ggbeeswarm)
library(ggpubr)
```

## Reading in the Data

```{r}
#READING IN THE DATA

#hospitalizations by county
hospitalizations <- read_csv(here("data", "asthma_hospitalizations.csv")) %>%
  clean_names()

#asthma prevalence by county
asthma_prevalence <- read_csv(here("data", "asthma_prevalence.csv")) %>%
  clean_names() 

#country demographic info
county_demographics <- read_csv(here("data", "county_demographics.csv")) %>%
  clean_names() 

#creating a list of the county yearly aqi data files 
aqi_files <- list.files(path = here("data"), pattern="annual_aqi*", full.names=TRUE, recursive=FALSE)

#looping through the files, reading them in and binding the rows to make one dataframe 
california_aqi <- aqi_files %>%
  map(function (path) {
    read_csv(path)
  }) %>%
  bind_rows() %>%
  clean_names() %>%
  filter(state =="California")
  
#county spatial data 
county_shapes <- read_sf(here("data", "CA_Counties", "CA_Counties_TIGER2016.shp")) %>%
  clean_names() 


#smoke pm2 data
smoke <- read_csv(here("data", "smoke_pm2_data.csv"))


```

## Cleaning the Data

```{r}
#CLEAN THE DATA

#filter to desired rows/columns from hospitalizations dataset
hosp_clean <- hospitalizations %>%
  filter(strata == "Total population",
         county != "California") %>%
  dplyr::select(county, year, number_of_hospitalizations, age_adjusted_hospitalization_rate) 



#filter to desired rows/columns from hospitalizations dataset
asthma_prev_clean <- asthma_prevalence %>%
  dplyr::select(county:comment) %>%
  filter(strata == "Total population",
         county != "California") %>%
  dplyr::select(county, years, current_prevalence) %>%
  mutate(current_prevalence = as.numeric(str_remove(current_prevalence, "%"))) #adjust percent variable to be numeric so it can be more easily used in analysis 
  

#filter rows, select columns, adjust columns, and create new columns in the demographics dataset. The race specific demographic info is an estimate for the overall 2010 to 2020 decade. 
county_dem_clean <- county_demographics %>%
  filter(state == "CA") %>%
  dplyr::select(county,
         ethnicities_american_indian_and_alaska_native_alone:ethnicities_white_alone_not_hispanic_or_latino,
         income_median_houseold_income, #median income 12 month median income collected from 2015 to 2019
         population_2020_population, population_population_per_square_mile) %>%
  rename("median_income" = "income_median_houseold_income",
         "pop2020" = "population_2020_population",
         "pop_density_sq_m" = "population_population_per_square_mile") %>%
  mutate(percent_poc = ethnicities_american_indian_and_alaska_native_alone + ethnicities_asian_alone + ethnicities_black_alone + ethnicities_hispanic_or_latino + ethnicities_native_hawaiian_and_other_pacific_islander_alone + ethnicities_two_or_more_races) %>% #For future analysis, I created a percent person of color variable to see whether asthma prevalence/hospitalizations was related to minority group status. 
  mutate(county = str_remove(county, " County")) %>%
  dplyr::select(!ethnicities_american_indian_and_alaska_native_alone:ethnicities_white_alone_not_hispanic_or_latino)


#select columns of interest and create new columns for aqi data
cali_aqi_clean <- california_aqi %>%
  dplyr::select(-state, -(days_co:days_pm10)) %>%
  mutate(good_or_moderate = good_days + moderate_days,
         unhealthy_or_worse = unhealthy_days + very_unhealthy_days + hazardous_days,
         unhealthy_or_worse_per365 = (unhealthy_or_worse / days_with_aqi)*365) 


#select column for geospatial data and rename county column to match other datasets
county_shape_clean <- county_shapes %>%
  dplyr::select(name) %>%
  rename("county" = "name")


smoke_clean <- smoke %>%
  mutate(GEOID = as.numeric(GEOID)) %>%
  filter(GEOID > 6000,
         GEOID < 6116) %>%
  mutate(GEOID = as.character(GEOID)) %>%
  mutate(GEOID = paste0("0", GEOID)) %>%
  mutate(date = lubridate::ymd(date)) %>%
  mutate(year(date))

smoke_clean_2015_to_2019 <- smoke_clean %>%
  filter(date > '2014-12-31',
         date < '2020-01-01') %>%
  rename("year" = "year(date)")

smoke_year_summaries <- smoke_clean_2015_to_2019 %>%
  group_by(GEOID, year) %>%
  summarize(mean_smokePM = mean(smokePM_pred),
            third_quartile_smoke_PM = quantile(smokePM_pred, 0.75),
            max_smoke_PM = max(smokePM_pred)) %>%
  left_join(county_shapes, by = c("GEOID" = "geoid")) %>%
  st_drop_geometry() %>%
  dplyr::select(GEOID, year, mean_smokePM, third_quartile_smoke_PM, 
                max_smoke_PM, name, namelsad) %>%
  rename("county" = "name")


```

## Joining Data

```{r}
# JOIN DATASETS FOR ANALYSIS

#add aqi and demographic data to the hospitalizations data set
hosp_full <- left_join(hosp_clean, cali_aqi_clean) %>%
  left_join(smoke_year_summaries) %>%
  left_join(county_dem_clean) %>%
  mutate(hosp_per_100k = (number_of_hospitalizations / pop2020) * 100000) %>%
  mutate(max_aqi = replace(max_aqi, max_aqi > 500, 500)) #since the EPA's AQI qualitative scale peaks at 500, I adjusted AQI values above 500 to just be 500 

```

## Checking Assumptions for Linear Regression

The hospitalization rate (per 100,000 people) will be used as the outcome variable for this model. It is important to note that an individual could be hospitalized more than once. Additionally, individuals may have traveled to a county and been hospitalized there, which would affect the

For each variable of interest, I created a basic plot to see whether the relationship looks linear, and transformed any variables that didn't. I then made a simple linear model using predictor of interest. Finally, I created a residual plot to make sure the errors were randomly distributed.

```{r}
##BASIC VISUALIZATION TO CHECK LINEAR REGRESSION ASSUMPTIONS


#median aqi on percent hosp rate
ggplot(data = hosp_full, aes(x = median_aqi, y = hosp_per_100k)) +
  geom_point()
#looks fairly linear but appears to be plateauing (or even starting to trend downwards) for the highest median aqi values. 

median_aqi_model <- lm(hosp_per_100k ~ median_aqi, data = hosp_full)

summary(median_aqi_model)

res <- resid(median_aqi_model)

plot(fitted(median_aqi_model), res) +
  abline(0,0)  #positive residuals are more extreme than negative residuals. It looks like the model is linear until about 62 median aqi, at which point it plateaus out. 



#log max aqi on percent hosp rate
ggplot(data = hosp_full, aes(x = log(max_aqi), y = hosp_per_100k)) +
  geom_point()

max_aqi_mod <- lm(hosp_per_100k ~ log(max_aqi), data = hosp_full)

summary(max_aqi_mod)

res <- resid(max_aqi_mod)

plot(fitted(max_aqi_mod), res) +
  abline(0,0) # appears fairly linear, although the residuals in the middle are more extreme than at the ends. Proceed with caution. 






# pop density on hosp rate
ggplot(data = hosp_full, aes(x = log(pop_density_sq_m), y = hosp_per_100k)) +
  geom_point()

pop_density_mod <- lm(hosp_per_100k ~ log(pop_density_sq_m), data = hosp_full)

res <- resid(pop_density_mod)

plot(fitted(pop_density_mod), res) +
  abline(0,0)



## median income on hosp
ggplot(data = hosp_full, aes(x = median_income, y = hosp_per_100k)) +
  geom_point()

income_mod <- lm(hosp_per_100k ~ median_income, data = hosp_per_100k)

res <- resid(income_mod)

plot(fitted(income_mod), res) +
  abline(0,0)  # a little bit of fanning ~ proceed with caution



## mean PM 2.5 from fire
ggplot(data = hosp_full, aes(x = sqrt(mean_smokePM), y = hosp_per_100k)) +
  geom_point()

smoke_mod <- lm(hosp_per_100k ~ sqrt(mean_smokePM), data = hosp_full)

summary(smoke_mod)

res <- resid(smoke_mod)

plot(fitted(smoke_mod), res) +
  abline(0,0)  # a little bit of fanning ~ proceed with caution



#unhealthy aqi days on percent hosp rate
ggplot(data = hosp_full, aes(x = unhealthy_or_worse_per365, y = hosp_per_100k)) +
  geom_point() # appears to be somewhat linear with a slope very close to 0. 

pct_bad_air_model <- lm(hosp_per_100k ~ unhealthy_or_worse_per365, data = hosp_full)

summary(pct_bad_air_model)

res <- resid(pct_bad_air_model)

plot(fitted(pct_bad_air_model), res) +
  abline(0,0) #most points are on the lower side for unhealthy or worse days per 365 days. Aside from that, residuals appear to have random distribution. Proceed with caution. 

```

## Deciding on a Model by Hand

```{r}
# Multiple regression models  

#largest model
summary(lm(hosp_per_100k ~ log(max_aqi) + median_income + unhealthy_or_worse_per365 + median_aqi + log(pop_density_sq_m) + sqrt(mean_smokePM) + max_smoke_PM, data = hosp_full))

#taking out unhealthy_or_worse_per265 and max_smoke_PM
summary(lm(hosp_per_100k ~ log(max_aqi) + median_income + median_aqi + log(pop_density_sq_m) + sqrt(mean_smokePM), data = hosp_full))

#taking out median_aqi as well
summary(lm(hosp_per_100k ~ log(max_aqi) + median_income + log(pop_density_sq_m) + sqrt(mean_smokePM), data = hosp_full))

#all variables were significant at this point. The effect of median_income appears quite minimal, but still highly significant. This model obtains the highest adjusted R-squared value as well. 
```

## Comparing my initial model selection to a model selection function

```{r}
# stepwise multiple regression model

null.lm <- lm(hosp_per_100k ~ 1, data = hosp_full)

full.lm <- lm(hosp_per_100k ~ log(max_aqi) + median_income + unhealthy_or_worse_per365 + median_aqi + log(pop_density_sq_m) + mean_smokePM + max_smoke_PM, data = hosp_full)

MASS::stepAIC(full.lm, scope = list(lower = null.lm, upper = full.lm), direction = "backward", k = log(60))  #Use backward elimination process to determine which variables are relevant

#                        Df Sum of Sq   RSS    AIC
# <none>                               52785 1354.2
# mean_smokePM           1    2245.3 55031 1360.5
# log(max_aqi)           1    5044.3 57830 1372.9
# log(pop_density_sq_m)  1    7675.0 60460 1384.0
# median_income          1    8856.1 61641 1388.8


```

## Checking Predictions 
```{r}
final_model <- lm(hosp_per_100k ~ log(max_aqi) + median_income + log(pop_density_sq_m) + sqrt(mean_smokePM), data = hosp_full)

predictions <- augment(final_model)

ggplot(data = predictions, mapping = aes(x = .fitted, y = hosp_per_100k)) +
  geom_point() +
  geom_abline(slope = 1, color = "red", lwd = 1) +
  xlim(20, 70) +
  annotate("text",
           x = 67, 
           y = 83, 
           label = "Line of perfect",
           color = "red",
           size = 3.5) +
  annotate("text",
           x = 67,
           y = 79,
           label = "predictions",
           color = "red",
           size = 3.5) +
  labs(x = "Predicted Hospitalizations (per 100k)",
       y = "True Hospitalizations (per 100k)",
       title = "Comparing our model predictions to the true value of hospitalizations",
       subtitle = "The model trends in the right direction, but tends to over predict low hospitalization rates and underpredict low hospitalization rates. A perfect model would follow the line in red.") +
  theme_classic()
```


## Joining the spatial data so spatial plots can be made

```{r}
## Joining the spatial data so spatial plots can be made

hosp_full_spatial <- hosp_full %>%
  left_join(county_shape_clean) %>%
  st_as_sf()

hosp_year_summaries <- hosp_full %>%
  group_by(county) %>%
  summarise(mean_hosp = mean(hosp_per_100k, na.rm = TRUE),
            mean_max_aqi = mean(max_aqi, na.rm = TRUE),
            mean_median_aqi = mean(median_aqi, na.rm = TRUE),
            mean_smoke = mean(mean_smokePM, na.rm = TRUE),
            max_smoke = max(max_smoke_PM, na.rm = TRUE)) %>%
  left_join(county_shape_clean) %>%
  st_as_sf()


asthma_prev_spatial <- asthma_prev_clean %>%
  left_join(county_shape_clean) %>%
  st_as_sf()

asthma_prev_spatial_2015_2016 <- asthma_prev_spatial %>%
  filter(years == "2015–2016")

asthma_prev_spatial_2017_2018 <- asthma_prev_spatial %>%
  filter(years == "2017–2018")
```

## Mapping Variables

```{r}
### Basic spatial plots

tm_shape(hosp_year_summaries) +
  tm_polygons(col = "mean_hosp",
              title = "Mean hospitalization rate (per 100,000)") +
  tm_layout(main.title = "Hospitalization Rate from Asthma between 2015–2019 for California Counies") 

tm_shape(hosp_year_summaries) +
  tm_polygons(col = "mean_median_aqi",
              title = "Median AQI") +
  tm_layout(main.title = "The Median AQI from 2015–2019 for California Counties") 
  

tm_shape(hosp_year_summaries) +
  tm_polygons(col = "mean_smoke",
              title = "Mean Smoke (in PM 2.5)") +
  tm_layout(main.title = "The Average Amount of Particulate Matter from Smoke between 2015–2019 for California Counties") 


tm_shape(asthma_prev_spatial_2015_2016) +
  tm_polygons(col = "current_prevalence")

tm_shape(asthma_prev_spatial_2017_2018) +
  tm_polygons(col = "current_prevalence")

```

$\alpha = 0.05$

$H_0:\mu_{hosp2018} - \mu_{hosp2016} = 0$

$H_A:\mu_{hosp2018} - \mu_{hosp2016} \neq 0$

```{r}
## Hypothesis test that hospitalizations would be different (likely higher) in 2018 than in 2016 due to increased fire acreage burnt. 

#2016: 669,534 burnt
#2018: 	1,975,086 burnt

hosp_2016_2018 <- hosp_full %>%
  filter(year == 2016 | year == 2018) %>%
  mutate(year = as.factor(year))
  
t.test(hosp_per_100k ~ year, data = hosp_2016_2018)


beeswarm_boxplot <- ggplot(data = hosp_2016_2018, mapping = aes(x = year, y = hosp_per_100k)) +
  geom_beeswarm() + 
  geom_boxplot(alpha = 0.3) +
  theme_classic() +
  labs(x = "Year",
       y = "Hospitalizations (per 100k)",
       title = "California County Hospitalization Rate from Asthma in 2016 and 2018",
       subtitle = "The distribution of hospitalization rates appear similar from 2016 and 2018")

beeswarm_boxplot + 
  stat_compare_means(method = "t.test", label.x = 1.35, label.y = 65)

```

My p-value of 0.85 indicates that based on this data, there is not enough evidence to conclude that there was a statistically significant difference in the mean hospitalization rate between 2016 and 2018.

However, it is important to note that this result does not mean that the total acreage of wildfires burnt has no effect on asthma hospitalizations. It's possible that the location of the fire and other factors, such as wind patterns and population density near the fire, are affecting the result. Further research could look into the spatial relationship between fires and where asthma related hospitalizations occur.






## Code Graveyard 

```{r}
# percent poc on percent hosp rate (maybe requires further transformation)
ggplot(data = hosp_full, aes(x = (percent_poc), y = hosp_per_100k)) +
  geom_point() +
  geom_smooth()

pct_poc_mod <- lm(hosp_per_100k ~ percent_poc, data = hosp_full)

summary(pct_poc_mod)

res <- resid(pct_poc_mod)

plot(fitted(pct_poc_mod), res) +
  abline(0,0)  #looks slightly exponential. Probably can't proceed. 
```

