---
title: "Asthma in California"
author: "Lewis White"
date: "2022-11-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
#LOADING PACKAGES
library(tidyverse)
library(lubridate)
library(janitor)
library(here)
library(purrr)
library(fs)
library(sf)
library(tmap)
```


```{r}
#READING IN THE DATA

#hospitalizations by county
hospitalizations <- read_csv(here("data", "asthma_hospitalizations.csv")) %>%
  clean_names() 

#asthma prevalence by county
asthma_prevalence <- read_csv(here("data", "asthma_prevalence.csv")) %>%
  clean_names() 

#country demographic info
county_demographics <- read_csv(here("data", "county_demographics.csv")) %>%
  clean_names() 


#creating a list of the county yearly aqi data files 
aqi_files <- list.files(path = here("data"), pattern="annual_aqi*", full.names=TRUE, recursive=FALSE)

#looping through the files, reading them in and binding the rows to make one dataframe 
california_aqi <- aqi_files %>%
  map(function (path) {
    read_csv(path)
  }) %>%
  bind_rows() %>%
  clean_names() %>%
  filter(state =="California")
  
#county spatial data 
county_shapes <- read_sf(here("data", "CA_Counties", "CA_Counties_TIGER2016.shp")) %>%
  clean_names() 

```

```{r}
#CLEAN THE DATA

#filter to desired rows/columns from hospitalizations dataset
hosp_clean <- hospitalizations %>%
  filter(strata == "Total population",
         county != "California") %>%
  select(county, year, number_of_hospitalizations, age_adjusted_hospitalization_rate)


#filter to desired rows/columns from hospitalizations dataset
asthma_prev_clean <- asthma_prevalence %>%
  select(county:comment) %>%
  filter(strata == "Total population",
         county != "California") %>%
  select(county, years, current_prevalence) %>%
  mutate(current_prevalence = as.numeric(str_remove(current_prevalence, "%"))) #adjust percent variable to be numeric so it can be more easily used in analysis 
  

#filter rows, select columns, adjust columns, and create new columns in the demographics dataset. The race specific demographic info is an estimate for the overall 2010 to 2020 decade. 
county_dem_clean <- county_demographics %>%
  filter(state == "CA") %>%
  select(county,
         ethnicities_american_indian_and_alaska_native_alone:ethnicities_white_alone_not_hispanic_or_latino,
         income_median_houseold_income, #median income 12 month median income collected from 2015 to 2019
         population_2020_population, population_population_per_square_mile) %>%
  rename("median_income" = "income_median_houseold_income",
         "pop2020" = "population_2020_population",
         "pop_density_sq_m" = "population_population_per_square_mile") %>%
  mutate(percent_poc = ethnicities_american_indian_and_alaska_native_alone + ethnicities_asian_alone + ethnicities_black_alone + ethnicities_hispanic_or_latino + ethnicities_native_hawaiian_and_other_pacific_islander_alone + ethnicities_two_or_more_races) %>% #For future analysis, I created a percent person of color variable to see whether asthma prevalence/hospitalizations was related to minority group status. 
  mutate(county = str_remove(county, " County")) %>%
  select(!ethnicities_american_indian_and_alaska_native_alone:ethnicities_white_alone_not_hispanic_or_latino)


#select columns of interest and create new columns for aqi data
cali_aqi_clean <- california_aqi %>%
  select(-state, -(days_co:days_pm10)) %>%
  mutate(good_or_moderate = good_days + moderate_days,
         unhealthy_or_worse = unhealthy_days + very_unhealthy_days + hazardous_days,
         percent_unhealthy_or_worse_days = (unhealthy_or_worse / days_with_aqi)*100) 


#select column for geospatial data and rename county column to match other datasets
county_shape_clean <- county_shapes %>%
  select(name) %>%
  rename("county" = "name")


#remove full data sets from environment to make calling the clean datasets easier 
rm(hospitalizations, asthma_prevalence, county_demographics, county_shapes, county_demographics)
```


```{r}
# JOIN DATASETS FOR ANALYSIS

#add aqi and demographic data to the hospitalizations data set
hosp_full <- left_join(hosp_clean, cali_aqi_clean) %>%
  left_join(county_dem_clean) %>%
  mutate(percent_hosp_rate = (number_of_hospitalizations / pop2020)*100) %>%
  mutate(max_aqi = replace(max_aqi, max_aqi > 500, 500)) #since the EPA's AQI qualitative scale peaks at 500, I adjusted AQI values above 500 to just be 500. 

```



```{r}
##BASIC VISUALIZATION TO CHECK LINEAR REGRESSION ASSUMPTIONS

# I will be using the percentage of people hospitalized for asthma as my outcome variable. It is important to note that an individual could be hospitalized more than once, so this does not imply that a unique percentage of the county was hospitalized for asthma related reasons. 

#for each variable of interest, I created a basic plot to see whether the relationship looks linear, and transformed any variables that didn't. I then made a simple linear model using predictor of interest. Finally, I created a residual plot to make sure the errors were randomly distributed. 



#unhealthy aqi days on percent hosp rate
ggplot(data = hosp_full, aes(x = percent_unhealthy_or_worse_days, y = percent_hosp_rate)) +
  geom_point()  

pct_bad_air_model <- lm(percent_hosp_rate ~ percent_unhealthy_or_worse_days, data = hosp_full)

summary(pct_bad_air_model)

res <- resid(pct_bad_air_model)

plot(fitted(pct_bad_air_model), res) +
  abline(0,0)


#median aqi on percent hosp rate
ggplot(data = hosp_full, aes(x = median_aqi, y = percent_hosp_rate)) +
  geom_point()  

median_aqi_model <- lm(percent_hosp_rate ~ median_aqi, data = hosp_full)

summary(median_aqi_model)

res <- resid(median_aqi_model)

plot(fitted(median_aqi_model), res) +
  abline(0,0)




#max aqi on percent hosp rate
ggplot(data = hosp_full, aes(x = log(max_aqi), y = percent_hosp_rate)) +
  geom_point()

max_aqi_mod <- lm(percent_hosp_rate ~ log(max_aqi), data = hosp_full)

res <- resid(max_aqi_mod)

plot(fitted(max_aqi_mod), res) +
  abline(0,0)



# percent poc on percent hosp rate (maybe requires further transformation)
ggplot(data = hosp_full, aes(x = percent_poc, y = percent_hosp_rate)) +
  geom_point()

pct_poc_mod <- lm(percent_hosp_rate ~ percent_poc, data = hosp_full)

res <- resid(pct_poc_mod)

plot(fitted(pct_poc_mod), res) +
  abline(0,0)



# pop density on hosp rate
ggplot(data = hosp_full, aes(x = log(pop_density_sq_m), y = percent_hosp_rate)) +
  geom_point()

pop_density_mod <- lm(percent_hosp_rate ~ log(pop_density_sq_m), data = hosp_full)

res <- resid(pop_density_mod)

plot(fitted(pop_density_mod), res) +
  abline(0,0)



## median income on hosp
ggplot(data = hosp_full, aes(x = median_income, y = percent_hosp_rate)) +
  geom_point()

income_mod <- lm(percent_hosp_rate ~ median_income, data = hosp_full)

res <- resid(income_mod)

plot(fitted(income_mod), res) +
  abline(0,0)


# max and median aqis to check for collinearity 
ggplot(data = hosp_full, aes(x = median_aqi, y = max_aqi)) +
  geom_point() 




# Multiple regression models  

summary(lm(percent_hosp_rate ~ log(max_aqi) + percent_poc + median_income + percent_unhealthy_or_worse_days + median_aqi + log(pop_density_sq_m), data = hosp_full))

summary(lm(percent_hosp_rate ~ log(max_aqi) + percent_poc + median_income + percent_unhealthy_or_worse_days + log(pop_density_sq_m), data = hosp_full))

summary(lm(percent_hosp_rate ~ log(max_aqi) + percent_poc + median_income + log(pop_density_sq_m), data = hosp_full))


# stepwise multiple regression model

null.lm <- lm(percent_hosp_rate ~ 1, data = hosp_full)

full.lm <- lm(percent_hosp_rate ~ log(max_aqi) + percent_poc + median_income + percent_unhealthy_or_worse_days + median_aqi + log(pop_density_sq_m), data = hosp_full)



library(MASS)

stepAIC(full.lm, scope = list(lower = null.lm, upper = full.lm), direction = "backward", k = log(60))  #Use backward elimination process to determine which variables are relevant


```


```{r}
## Joining the spatial data so spatial plots can be made

hosp_full_spatial <- hosp_full %>%
  left_join(county_shape_clean) %>%
  st_as_sf()

hosp_year_summaries <- hosp_full %>%
  group_by(county) %>%
  summarise(mean_hosp = mean(percent_hosp_rate, na.rm = TRUE),
            mean_max_aqi = mean(max_aqi, na.rm = TRUE),
            mean_median_aqi = mean(median_aqi, na.rm = TRUE)) %>%
  left_join(county_shape_clean) %>%
  st_as_sf()


asthma_prev_spatial <- asthma_prev_clean %>%
  left_join(county_shape_clean) %>%
  st_as_sf()

asthma_prev_spatial_2015_2016 <- asthma_prev_spatial %>%
  filter(years == "2015–2016")

asthma_prev_spatial_2017_2018 <- asthma_prev_spatial %>%
  filter(years == "2017–2018")
```


```{r}
### Basic spatial plots

tm_shape(hosp_year_summaries) +
  tm_polygons(col = "mean_hosp")

tm_shape(hosp_year_summaries) +
  tm_polygons(col = "mean_max_aqi")

tm_shape(hosp_year_summaries) +
  tm_polygons(col = "mean_median_aqi")

tm_shape(asthma_prev_spatial_2015_2016) +
  tm_polygons(col = "current_prevalence")

tm_shape(asthma_prev_spatial_2017_2018) +
  tm_polygons(col = "current_prevalence")

```





$\alpha = 0.05$

$H_0:\mu_{hosp2018} - \mu_{hosp2016} = 0$

$H_A:\mu_{hosp2018} - \mu_{hosp2016} \neq 0$

```{r}
## Hypothesis test that hospitalizations would be different (likely higher) in 2018 than in 2016 due to increased fire acreage burnt. 

#2016: 669,534 burnt
#2018: 	1,975,086 burnt

hosp_2016_2018 <- hosp_full %>%
  filter(year == 2016 | year == 2018)
  
t.test(percent_hosp_rate ~ year, data = hosp_2016_2018)
```

My p-value of 0.85 indicates that based on this data, there is not enough evidence to conclude that there was a statistically significant difference in the mean hospitalization rate between 2016 and 2018. 

However, it is important to note that this result does not mean that the total acreage of wildfires burnt has no effect on asthma hospitalizations. It's possible that the location of the fire and other factors, such as wind patterns and population density near the fire, are affecting the result. Further research could look into the spatial relationship between fires and where asthma related hospitalizations occur. 






